---
title: "multChernoff: Get Started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{multChernoff: Get Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
library(multChernoff)
```

# Overview

`multChernoff` computes finite-sample upper bounds for the likelihood ratio
test (LRT) under multinomial sampling. These bounds can be used to obtain
conservative p-values and critical values when the sample size is not large
relative to the alphabet size, so standard chi-squared asymptotics may be
inaccurate.

The package exposes three main functions:

- `mgfBound()` evaluates the upper bound on the moment generating function.
- `tailProbBound()` returns a conservative upper bound on `P(LRT > x)`.
- `criticalValue()` returns a finite-sample critical value for a target tail
  probability.

# When to Use This Package

The package is most useful when:

- the data are multinomial or a product of independent multinomial samples
- the alphabet size is not small relative to the sample size
- asymptotic chi-squared calibration may be too optimistic

In this setting, `tailProbBound()` and `criticalValue()` provide conservative
finite-sample alternatives to asymptotic likelihood-ratio approximations.

# A Simple Tail Bound

The most direct workflow is to evaluate the finite-sample tail bound for a
fixed LRT value.

```{r}
tailProbBound(x = 20, k = 7, n = 50)
```

This value is a conservative upper bound on the tail probability. For
comparison, the package also supports independent multinomial samples by
passing vectors of matching lengths for `k` and `n`.

```{r}
tailProbBound(x = 12, k = c(3, 4), n = c(12, 20))
```

# Computing a Critical Value

To calibrate a test at level `p`, use `criticalValue()`.

```{r}
criticalValue(k = 10, n = 40, p = 0.05)
```

The returned value can be used as a finite-sample alternative to the standard
chi-squared cutoff.

For comparison, here is the usual asymptotic cutoff for the same nominal level.

```{r}
qchisq(0.95, df = 9)
```

# Inspecting the MGF Bound

For lower-level diagnostics, `mgfBound()` evaluates the bound on the moment
generating function at a given value of `lambda`.

```{r}
mgfBound(k = 10, n = 20, lambda = c(0, 0.25, 0.5))
```

# Workflow Summary

In practice, the package workflow is:

1. Compute an observed likelihood-ratio statistic from multinomial data.
2. Use `tailProbBound()` to obtain a conservative p-value upper bound.
3. Or use `criticalValue()` to calibrate a finite-sample rejection threshold.
4. Use the resulting bound or critical value inside downstream inference or
   optimization problems.

# Relation to the README

The README is best used as the landing page for the package website: short
motivation, installation, and a minimal example. This vignette is the compact
"Get started" article that explains the core API. Longer applications, such as
the Corbet butterfly example, should live in separate future articles.

# References

F. Richard Guo and Thomas S. Richardson (2021). Chernoff-Type Concentration of
Empirical Probabilities in Relative Entropy. *IEEE Transactions on Information
Theory*, 67(1), 549-558. <https://doi.org/10.1109/TIT.2020.3034539>
